---
title: "Overview"
format: 
  html:
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(knitr)
library(data.table, quietly=T, verbose =F)
library(dplyr, quietly=T, verbose =F)
library(reshape2, quietly=T, verbose =F)
library(sf, quietly=T, verbose =F)
library(mapsf, quietly=T, verbose =F)
```

## Introduction

We present here some preliminary experiments in order to give an overview of the possibilities of analysis of world region based on the Food Balance Sheets (FBS) published by FAO.

## Sources

We use in a first step the most recent data produced by FAO, covering the period 2010 to present. We will further used another database covering the period 1963-2013 but not immediately as they are some difficulties for the harmonization of the two databases.

### Geometry

We have prepared different geometry adapted to the 188 countries or territories available in the database.

```{r}
map<-readRDS("geom/wld_all.RDS")
map3<-readRDS("geom/wld_states_fao.RDS")
map2<-readRDS("geom/wld_region_fao.RDS")
map1<-readRDS("geom/wld_continent_fao.RDS")
```

```{r}

mf_map(map1, type="typo", var="continent")
mf_map(map, type="base", col=NA, add=T, lwd=0.6)
mf_map(map2, type="base", col=NA, add=T, lwd=2)
mf_layout("179 states located in 5 continents and 22 regions", frame=T)
```

### Population

The FAO has collected variables from other sources (UN, Worldbank,...) in order to get estimates of the population of territorial units for which the FBS has been collected. We eliminate the unit called *China* (F351) because it is the sum of four other units present in the database *China Mainland* (CHN), *Hong Kong* (HKG), *Macao* (MAC) and ... *Taïwan* (TWN). This duplication of data is clearly not practical from statistical point of view. But we can easily imagine why the republic of China has obliged the FAO to proceed this way ...

```{r}
pop<-fread("source/fbs2010/FAOSTAT_data_en_11-13-2024_POP.csv") 
pop <- pop %>% select(iso3 = `Area Code (ISO3)`, year=Year, pop=Value) %>%
         filter(iso3 !="F351", iso3 %in% map3$iso3) 

pop<-pop %>% left_join(st_drop_geometry(map3)) %>% as.data.frame()

kable(head(pop), caption = "Population 2010-2022 - Long format")
```

The dataset is in long format but can easily be transformed in wide format :

```{r}
x<-  reshape2::dcast(pop, formula = iso3~year, value.var="pop",fill = NA)
kable(head(x), caption = "Population 2010-2022 - wide format")

```

Some missing values can be observed but only for a limited number of territorial units :

```{r}
mis <- pop %>% 
        group_by(iso3, name) %>% 
       summarise(available = n(), max = 13 ,pct = round(100*n()/13,1)) %>%
        filter(pct<100) %>% arrange(pct)
kable(mis)
```

We elaborate a map of the average share of world population over the period :

```{r}
avgpop<-pop %>% group_by(iso3) %>% summarise(popmean = mean(pop, na.rm=T)) %>%
         mutate(pct = 100*popmean / sum(popmean))

mapdon <-map3 %>% left_join(avgpop)

mf_map(map, type="base", col="lightyellow")
mf_map(mapdon, type="prop",var="pct", leg_title = "%", inches =0.1)
mf_layout(title = "Average share of world population 2010-2022",
          credits = "FAO - Food Balance Sheets - 2010-2022",
          frame=T)
```

### Food Balance Sheet (Kcal)

We select the Food Balance Sheet (FBS) measured in Kcal/capita/day which is one of the four possible option of measure. Initialy the table has a lot of columns :

#### Structure of data

```{r}
x <- fread("source/fbs2010/FAOSTAT_data_en_11-13-2024-KCAL.csv") 
x<-x %>% filter(`Area Code (ISO3)` != "F351",
                `Item Code (FBS)`!="S2905",
                `Item Code (FBS)`!="S2907",)
kable(head(x,10), caption = "FBS 2010-2022 - Raw data")

```

We decide to simplify a bit the structure and to reduce the number of columns like this :

```{r}
fbs<-x %>% select(year= Year,iso3=`Area Code (ISO3)`, item = `Item Code (FBS)`, item_name=Item, K=Value) %>% 
  filter(iso3 != "F351",  # Eliminate China aggregated
         item!="S2905",   # Eliminate aggregate of cereals   
         item!="S2907")    # Eliminate aggregate of Starchy Roots
kable(head(fbs,10), caption = "FBS 2010-2022 - Simplified")
```

#### Hierarchisation of food items

In the initial format delivered by FAO, the classification of items is a mixture of aggregates at different hierarchical level. Here we have decided to collect only the lowest level and to add a hierarchy of aggregation levels based on the table below :

```{r}
code_items<-read.csv2("data/FBS_SUA_codes_corrected.csv",encoding = "UTF-8") %>% select(1:8) %>% unique()
names(code_items)<-c("i0","name_i0","i1","name_i1","i2","name_i2","i","name_i")
kable(code_items,caption = "Hierarchy of food items")

### CHECK
# w<- x %>% select(`Item Code (FBS)`,Item) %>% unique()
# names(w) <-c("i","name_base")
# w$i<-substr(w$i,2,5)
# test2<-code_items[,7:8]
# test2$i<-as.character(test2$i)
# check = full_join(w,test2)

```

### Synthetic data base

We decide to add the hiearachy of food items but also the hierarchy of geographical levels and the population in a single file where it will be possible to realize easily all types of analysis.

```{r}
names(fbs)<-c("t","s","i","name_i","Kist")
fbs$i<-as.numeric(substr(fbs$i,2,5)) 
code_items<-code_items %>% select(-name_i)
base <- fbs %>% left_join(code_items) %>% filter(is.na(i0)==F)
names(pop)<-c("s","t","Pst","name_s","reg_s","cont_s")
pop$t<-as.numeric(pop$t)
base <- base %>% left_join(pop)
base<-base[,c(2,1,13,3,4,5,12,6:11,14:15)]
kable(head(base),caption = "Structure of multilevel database")
saveRDS(base,"data/base_fao_Kcal_2010_2022.RDS")
class(base)
```

As we can see the database is a "cube" with three elementary dimensions :

-   $t$ : the year of observation
-   $s$ : the state concerned
-   $i$ : the item of food at the elementary level.

Two quantitative indicators are available :

-   $F_{ijt}$ : the consumption of food from item $i$ in state $s$ during the year $t$ measured in kcal/capita/day
-   $P_{it}$ : the estimated population of state $i$ during the year $t$ which can be used as weighting criteria for the procedure of agregation if we want to obtain corrected value at world, continental or regional level.

The following columns are different keys of aggregation for food items (i0, i1,i2) or for spatial units (reg, cont)

## Test of the database

The synthetic database has 171069 lines which is less than the total number of value of the cube that should be equal to 83 items x 190 countries x 13 years = 205010. It means that the matrix is filled at 83.4% but 16.6% of values are missing (no data for some countries some years) or equal to 0 (items that are not reported in one country).

### Global Food Balance in 2021

We can firstly try to compute some global values, for example in 2021.

#### World level

```{r}
base <-readRDS("data/base_fao_Kcal_2010_2022.RDS")
x<-base %>% filter(t=="2021") %>%
             group_by(s) %>%
            summarise(K=sum(Kist),
                      P = max(Pst)) %>%
            mutate(KP=K*P) %>%
           summarise(KP=sum(KP,na.rm=T),P=sum(P,na.rm=T))%>%
           mutate(K=KP/P)
kable(x, caption = "Food ration in Kcal by day at world level in 2021",col.names = c("Total Kcal","Total Population", "Kcal/capita")
      )
```

- **Comment** : If our results are correct, the average food ration at world level (weighted by population of countries) is equal to **2978 kcal/capita/day**. This value is based on 190 coutries that summarize a total of 7.850 billions of inhabitants. We have verified in the publication of FAO that it is exactly the figure published in the report of the  organization.



#### Continental level

```{r}
base <-readRDS("data/base_fao_Kcal_2010_2022.RDS")
x<-base %>% filter(t=="2021", is.na(cont_s)==F) %>%
             group_by(s,cont_s) %>%
            summarise(K=sum(Kist),
                      P = max(Pst)) %>%
            mutate(KP=K*P) %>%
            group_by(cont_s)%>%
           summarise(KP=sum(KP,na.rm=T),P=sum(P,na.rm=T))%>%
           mutate( KPpct=100*KP/sum(KP),Ppct=100*P/sum(P), K=KP/P, Index=100*KPpct/Ppct) %>% 
  arrange(Index)
kable(x, caption = "Food ration in Kcal by day at continental level in 2021",
      col.names = c("Continent","Kcal","Population","% Kcal","% Pop", "Kcal/capita", "index 100 = World"),
      digits=c(0,0,0,2,2,0)
      )
```

- **Comment** : At continental level, we add some columns in order to compare for each continent the share of population and the share of food consumption measured in Kcal. According to our results, the lowest level of consumption is observed in **Africa** which account for 17.7% of world population but only 15.3% of food consumption mesured in Kcal. The level of consumption of Africa is at index 86 i.e -14% lower than the world average. On the contrary, Europe is at index 116 which is +16% higher than the world average. As inprevious case, we obtain the same figure than the FAO report. 



#### Regional level

```{r}
base <-readRDS("data/base_fao_Kcal_2010_2022.RDS")
x<-base %>% filter(t=="2021", is.na(cont_s)==F) %>%
             group_by(s,reg_s) %>%
            summarise(K=sum(Kist),
                      P = max(Pst)) %>%
            mutate(KP=K*P) %>%
            group_by(reg_s)%>%
           summarise(KP=sum(KP,na.rm=T),P=sum(P,na.rm=T))%>%
           mutate( KPpct=100*KP/sum(KP),Ppct=100*P/sum(P), K=KP/P, Index=100*KPpct/Ppct) %>% arrange(Index)
kable(x, caption = "Food ration in Kcal by day at regional level in 2021",
      col.names = c("region","Kcal","Population","% Kcal","% Pop", "Kcal/capita", "index 100 = World"),
      digits=c(0,0,0,2,2,0)
      )
```

- **Comment** : We can replicate the same analysis at the level of the 22 regions defined by UN as subdivisions of continents. We can reveal more important inequalities with a lowest level of 76 (-24% of world average) in Eastern Africa and an highest level of 130 (+30% of world average) in Northern America. These results are consistent with the publication of the FAO and the litterature on the subject.






